Comparing software variability across and within fMRI analysis packages

### 2022-04-12 paper/paper.tex
Minor revisions

\maketitle
\begin{abstract}
  Variability has been broadly observed in functional MRI analyses as a
  result of software differences both between and within analytic tools.
  However, the relationship between within-tool and between-tool software
  variabilities for a given analysis is unclear. Within-tool variability has multiple origins
  including variations in algorithms, parametrizations, and
  low-level software dependencies. We focus on the effects resulting from software dependencies --- 
  in particular due to operating system and other infrastructural updates --- as they often remain unnoticed 
  while algorithms and parametrizations are more controllable.
 We extended
  a previous comparison of fMRI analysis software libraries (namely FSL,
  AFNI, and SPM) and related the observed differences to within-tool
  software variability simulated using Monte-Carlo arithmetic. In group
  analyses, we found that
  between-tool software variability was consistently larger than within-tool software
  variability. In subject analyses, within-tool software variability
  approached between-tool software variability in some regions for some subjects.
  Interestingly, the brain masking instability was triggered by both within- and between-tool
  software variabilities.
  % Finally, we found that
  % within-tool software variability had effects of similar magnitude to between-tool
  % software variability when simulated from numerical perturbations introduced at the virtual
  % precision of 17 bits --- a high but plausible level of uncertainty for
  % scientific computing libraries. \tristan{Revise from latest results}
  % %  \gk{What does this mean? They only have
  % precision of 17, or they use data that is 17 bits?}
  %  \yohan{agree with Greg, it is unclear and a strong assertion!}
  %  \camille{It is not entirely clear to me how this relates to the previous
  %  conclusions that BT variability was consistently larger than machine
  %  error} \tristan{Better?}
  Our findings motivate the continued investigation of within-tool
  software variability in neuroimaging, and position it as a possible proxy
  for studying some of the arbitrary between-tool software variations for
  subject-level analyses.
\end{abstract}

@@

We verified the correctness of our analyses by comparing our unperturbed
t-statistic group maps with the ones obtained
in~\cite{bowring2019exploring}. For FSL, we found
strictly identical results with the same MD5 checksums. For SPM, the files (as measured by MD5 checksums) were different but differences
were visually unnoticeable. For AFNI, the overall patterns of activations were preseved
, however, differences were noticeable visually.
The observed differences remained small (see Supplemental Material~\ref{sec:supp-repro}), and might be due to the use of
GNU/Octave vs MATLAB in SPM, and of multithreading in AFNI. We performed visual quality control of the AFNI
and SPM results for each individual subject and confirmed that T1-weighted images were
correctly skull-stripped and registered to the MNI template.

@@

Table~\ref{table:pipeline-stats} presents summary statistics for variability in the
group-level t-statistic.
Mean of differences in t-statistic maps was centered around zero for BT and WT
in both thresholded and unthresholded maps (t-test p~\textless~$10^{-5}$).
However, for each tool pair (A, B), BT variability was
significantly larger than WT in tool A or B,
with an order of magnitude larger standard deviation of differences 
in both thresholded and unthresholded maps, 
(Wilcoxon signed-rank test and t-test p~\textless~$10^{-5}$).
These global differences were confirmed by Bland-Altman plots showing a
clear dominance of BT over WT
(Figure~\ref{fig:unthresh-maps}-\textbf{A},\textbf{B}).
In addition, results show that BT and WT were not correlated across voxels
(Pearson's r $\approx 0$, p$< 10^{-5}$, Figure~\ref{fig:unthresh-maps-corr}).
% Interestingly, the correlation appears driven by a set of voxels exhibiting comparable BT and
% WT values located around the identity line in Figure~\ref{fig:unthresh-maps}D, which, however, did not seem to have
% any spatial consistency.

@@

While the previous results were obtained over the union of t-statistic maps between the pair of tools,
we also computed results in the intersection of maps and found that the brain masking instability was
triggered by both BT and WT.
BT and WT variabilities were consistently smaller than ones with the fixed masks
in both thresholded and unthresholded maps (see Table~\ref{table:pipeline-stats-fixed-mask}).
The effect of masking appeared in Bland-Altman plots by showing that blob lines
were removed after fixing masks (Figure~\ref{fig:unthresh-maps-mask}-\textbf{A},\textbf{B}).

@@

For both group- and subject-level analyses, between- and within-tool
software variabilities similarly trigger the brain masking instability.
Even though both types of variability are different in nature, this result suggests
that in some cases they may have a common cause that might be related to
the conditioning of fMRI analysis in a specific dataset. 
% Along these lines,
% in some regions, instabilities of similar magnitude may be triggered by
% small numerical perturbations, model variations, or implementation
% differences. For instance, high-motion datasets may trigger between- and
% within-tool instabilities similarly.

Therefore, numerical stability may be a suitable proxy to study
between-tool variability. This speculation might be of practical value to
address software variability at large given that within-tool variability
measurements can be automated while between-tool ones require substantial
human intervention. Indeed, within-tool variability can be simulated
irrespective of the particular analysis done, and with important sample
sizes, while between-tool variability requires re-implementing analysis in
different toolboxes and is limited to a few samples. Automated multi-verse
frameworks could be developed to assess and possibly reduce within-tool
variability as a pre-requisite for between-tool variability.

% \tristan{This paragraph needs to be revised as we can't conclude that from the data we have}
% The finding that numerical variability approached tool variability at the
% virtual precision of t=17~bits is interesting too. Indeed, while machine
% error generally introduces perturbations on the least-significant bit of the mantissa,
% ones 
% % \yohan{I would not mix virtual precision and error measurement}\tristan{better?}
% common scientific software dependencies introduce larger perturbations
% .
% For instance, SciPy's 2D spline interpolation was recently found to be
% precise up to 10~bits~\cite{pytracer}
% and its replacement by a more precise implementation might therefore introduce
% numerical perturbations leading to errors in the range of between-tool
% variability. Such high errors
% might be triggered by updates in operating systems, Python, MATLAB, and
% other software dependencies.
% \yohan{The sentence is ok, just a picky point: the SciPy result you cited
% describes the precision of the spline implementation while at the end
% you're talking differences due to updates. The nuance is that two versions
% may have a similar precision and so when you udpate your system, the
% difference between both versions is low but each version can be
% ill-conditionned.}
% \tristan{better?}

@@

In conclusion, our results motivate further numerical stability
investigations in fMRI analyses. Pipeline-level analyses could be conducted
to identify specific components that contribute to numerical variability,
and if possible correct them accordingly. Besides, in cases where
instability is inherent to the analysis, sampling results distributions
through numerical perturbations might improve stability, as explored
in~\cite{kiar2021data}. Finally, pipeline-specific statistical corrections
might be envisaged to account for within-tool software variability.

>>>>

\maketitle
\begin{abstract}
  Variability has been broadly observed in functional MRI analyses as a
  result of software differences both between and within analytic tools.
  However, the relationship between within-tool and between-tool software
  variabilities for a given analysis is unclear. Within-tool variability has
  multiple origins including variations in algorithms, parametrizations, and
  low-level software dependencies. We focus on the effects resulting from
  software dependencies --- in particular due to operating system and other
  infrastructural updates --- as they often remain unnoticed while algorithms
  and parametrizations are more controllable. We extended a previous
  comparison of fMRI analysis software libraries (namely FSL, AFNI, and SPM)
  and related the observed differences to within-tool software variability
  simulated using Monte-Carlo arithmetic. In group analyses, we found that
  between-tool software variability was consistently larger than within-tool
  software variability. In subject analyses, between-tool software
  variability also dominated within-tool variability overall, however,
  within-tool software variability approached between-tool software
  variability in some regions for some subjects. Interestingly, the brain
  masking instability was triggered by both within- and between-tool software
  variabilities. Our findings motivate the continued investigation of
  within-tool software variability in neuroimaging.
  \end{abstract}

@@

We verified the correctness of our analyses by comparing our unperturbed
t-statistic group maps with the ones obtained
in~\cite{bowring2019exploring}. For FSL, we found
strictly identical results with the same MD5 checksums. For SPM, the files (as measured by MD5 checksums) were different but differences
were visually unnoticeable. For AFNI, the overall patterns of activations were preserved, however, differences were noticeable visually.
The observed differences remained small (see Supplemental Material~\ref{sec:supp-repro}), and might be due to the use of
GNU/Octave vs MATLAB in SPM, and of multithreading in AFNI. We performed visual quality control of the AFNI
and SPM results for each individual subject and confirmed that T1-weighted images were
correctly skull-stripped and registered to the MNI template.

@@

Table~\ref{table:pipeline-stats} presents summary statistics for variability in the
group-level t-statistic.
Mean of differences in t-statistic maps was centered around zero for BT and WT
in both thresholded and unthresholded maps (t-test p~\textless~$10^{-5}$).
However, for each tool pair (A, B), BT variability was
significantly larger than WT in tool A or B
(Wilcoxon signed-rank test and t-test p~\textless~$10^{-5}$).
These global differences were confirmed by Bland-Altman plots showing a
clear dominance of BT over WT
(Figure~\ref{fig:unthresh-maps}-\textbf{A},\textbf{B}).
In addition, results show that BT and WT were not correlated across voxels
(Pearson's r $\approx 0$, p$< 10^{-5}$, Figure~\ref{fig:unthresh-maps-corr}).
% Interestingly, the correlation appears driven by a set of voxels exhibiting comparable BT and
% WT values located around the identity line in Figure~\ref{fig:unthresh-maps}D, which, however, did not seem to have
% any spatial consistency.

@@

While the previous results were obtained over the union of t-statistic maps between the pair of tools,
we also computed results in the intersection of maps and found that the brain masking instability was
triggered by both BT and WT (Table~\ref{table:pipeline-stats-fixed-mask}).
BT and WT variabilities were generally smaller than the ones obtained previously (Table~\ref{table:pipeline-stats})
in both thresholded and unthresholded maps.
The effect of masking appeared in Bland-Altman plots by showing that blob lines
were removed after fixing masks (Figure~\ref{fig:unthresh-maps-mask}-\textbf{A},\textbf{B}).

@@

For both group- and subject-level analyses, between- and within-tool
software variabilities similarly trigger the brain masking instability.
Even though both types of variability are different in nature, this result
suggests that in some cases they may have a common cause that might be
related to the conditioning of fMRI analysis in a specific dataset. In such
cases, numerical stability may be a suitable proxy to study between-tool
variability, which could be of practical value.

@@

In conclusion, between-tool variability clearly dominates within-tool
variability overall. However, within-tool variability still reaches
magnitudes that may affect scientific conclusions derived from data
analyses --- in particular in subject analyses and in group analyses
that rely on thresholded maps. Moreover, sources of variability such as brain
masking affect both within-tool and between-tool
variability: in these specific cases only, within-tool variability may be
used as a proxy to between-tool variability. These findings motivate
further investigations of within-tool varability in fMRI analyses.
